Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-02-26 Dail_Mad_ext.tex
Several minor edits2

\author{Richard B. Chandler  \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708
   \and Jeffrey A. Hostetler \footnote{Contact:
     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
%        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012}
\maketitle

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
In the Poisson case, the equivalent expression of
$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
\begin{equation}
[N_{i1} | \Lambda] =
\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
  \label{eq:N1br}
\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpreted as detection
probability parameters.

@@

Maximum likelihood estimation of parameters in models with
discrete random effects requires removing them by
summation. For Markovian models, this can be done recursively,
making the likelihood tractable. A general expression of the likelihood
can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $K$, which
%%% Can we rename the upper bound?  We're also using K
%%% for equilibrium abundance.
should be high enough such that the MLEs are not affected by increasing
it further. This likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

@@

Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data. [EXPLAIN POSTERIORS AND PRIORS]
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. Although MCMC can be slow, our experience is that it
may actually converge faster than the amount of time required to
maximize the likelihood shown in Eq.\ref{eq:like}, especially
when the upper bound of the summation must be high, e.g. $K>200$.
Furthermore, MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing
users to specify the model using a simple symbolic descriptions. For
example here is \textbf{BUGS} code for the ``geometric'' model described
in Eq.~\ref{eq:hm}:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma*N[i,t-1])
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,t])
    }
  }
}
\end{verbatim}

@@

Variation in detection probability among observers is another
source of random variation that may need to be accounted for applying
the DM model. For example, differences in observers' ability to see,
hear, or identify birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS 
\citep{robbins_etal:1986,diefenbach_etal:2003,sauer_etal:1994auk,alldredge_etal:2007auk,campbell_francis:2011}.
Estimating a separate detection probability for each observer
can be difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

Current BBS trend estimators deal with this problem by
treating observer identity as a random (as opposed to a fixed)
effect \citep{link_sauer:2002,sauer_link:2011}.
%(Link and Sauer 2002, Sauer and Link 2011).
This allows observer-specific differences in detection probability,
but assumes that observers are selected at random from a pool
of potential observers.  Models that contain both random and
fixed effects are referred to as mixed models.  Often in mixed
models (as in this case), the random effect is modeled not
because it of interest in itself, but to avoid bias in the
estimates of the fixed effects.

To include random observer effects in DM models and these
extensions, Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
X_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $X_{ijt}$ is the number of bird recorded at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the standard deviation of the random observer effects (also on the logit scale). 
We have implemented random observer effects in the Bayesian
framework using program \textbf{JAGS} \citep[version 3.2.0]{plummer:2003}
% (version 3.2.0; Plummer 2003)
with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.

@@

The classical approach to computing $[N^+_t]$ is based upon
Empirical Bayes methods. Note that neither Empiriis not actually a
Bayesian method, it is used to estim.
We take the parameter estimates and plug them back in.... For
example to obtain the posterior distribution for $[N_{i1}]$
\begin{equation}
[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
\hat{p}]}{\sum_{N_{i1}=0}^K
[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
  \label{eq:eb1}
\end{equation}
Once we have estimated the posterior distribution for the first
time period, we can then proceed to estimate the posterior
distribution for subsequent time periods using recursion.
\begin{equation}
  [\hat{N}_{it}] =
  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
  \label{eq:eb2}
\end{equation}

Note that when $X_{it}$ is not observed, including for future
time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
uniform such that it has no influence on the posterior. In other words,
the posterior for years with no observed data is determined entirely
by the estimated posterior from the previous year and the
population dynamics parameters.

If covariates of the dynamics parameters are present, the problem
becomes much harder because, to fully account for uncertainty,
we need a model for the future values of the covariates as well. When
the covariates are continuous, this requires integrating over their
possible values, which becomes very challenging
computationally. Indeed, simulation methods such as MCMC are the
only viable options when numerous covariates are present, which will
often be the case in many ecological studies.

@@

We estimated the parameters for each simulation using the same
initial abundance (Poisson) and dynamics models as were
simulated, implemented in the unmarked library in \textbf{R}.
When run in a maximum likelihood framework, these models require a
maximum abundance to integrate over
\citep{royle:2004biom,dail_madsen:2011};
%(Royle 2004, Dail and Madsen 2011);
we used 200.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95\% confidence
intervals for parameters that overlap the true values).

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each
species, and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant, autoregressive, Ricker-logistic, Gompertz-logistic, autoregressive +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a maximum abundance to integrate
over when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and, where appropriate, zero-inflated dynamics.  We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 tuning samples.

@@

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
classical and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}

>>>>

\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
%        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{RChandler@usgs.gov}, +001-301-497-5856} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

Recall that conventional state-space models assume that the
distribution for the initial time period,
$[N_1 | \bm{\theta}]$, is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
i.e. the model for abundance dynamics. This state
model has allowed us to replace the initial abundance
distribution and transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and
(3) have parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be
more easily interpreted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
Under this model, $\bm{p}$ is interpreted as the detection
probability parameter.

@@

Maximum likelihood estimation of parameters in models with
discrete random effects requires removing them by
summation. For Markovian models, this can be done recursively,
making the likelihood tractable. A general expression of the likelihood
can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $N_{max}$, which
%%% Can we rename the upper bound?  We're also using K
%%% for equilibrium abundance.
should be high enough such that the MLEs are not affected by increasing
it further. This likelihood can be maximized numerically using the
\textbf{R} package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples
are given in the Appendix.

@@

Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data. [EXPLAIN POSTERIORS AND PRIORS]
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. Although MCMC can be slow, our experience is that it
may actually converge faster than the amount of time required to
maximize the likelihood shown in Eq.\ref{eq:like}, especially
when the upper bound of the summation must be high, e.g. $N_{max}>200$.
Furthermore, MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing
users to specify the model using a simple symbolic descriptions.

%For example here is \textbf{BUGS} code for the ``geometric'' model described
%in Eq.~\ref{eq:hm}:
%\begin{verbatim}
%model {
%lambda ~ dunif(0, 5)
%omega ~ dunif(0,1)
%gamma ~ dunif(0, 10)
%p ~ dunif(0,1)
%for(i in 1:nSites) {
%  N[i,1] ~ dpois(lambda)
%  y[i,1] ~ dbin(p, N[i,1])
%  for(t in 2:nYears) {
%    S[i,t-1] ~ dbin(omega, N[i,t-1])
%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
%    N[i,t] <- S[i,t-1] + G[i,t-1]
%    n[i,t] ~ dbin(p, N[i,t])
%    }
%  }
%}
%\end{verbatim}

@@

Variation in detection probability among observers is another
source of random variation that may need to be accounted for applying
the DM model. For example, differences in observers' ability to see,
hear, or identify birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS 
\citep{robbins_etal:1986,sauer_etal:1994auk,diefenbach_etal:2003,campbell_francis:2011}.%,alldredge_etal:2007auk
%Estimating a separate detection probability for each observer
%can be difficult and reduces one's ability to estimate the quantities
%of interest.  This problem is compounded by the fact that
%observers differ greatly in the number of surveys they have
%run (so that many observers' separate detection probabilities
%could not be accurately estimated).

Current BBS trend estimators deal with this problem by
treating observer identity as a random %(as opposed to a fixed)
effect \citep{link_sauer:2002,sauer_link:2011}.
%(Link and Sauer 2002, Sauer and Link 2011).
%This allows observer-specific differences in detection probability,
%but assumes that observers are selected at random from a pool
%of potential observers.  Models that contain both random and
%fixed effects are referred to as mixed models.  Often in mixed
%models (as in this case), the random effect is modeled not
%because it of interest in itself, but to avoid bias in the
%estimates of the fixed effects.
%
To include random observer effects in DM models and these
extensions, Eq.~\ref{eq:p1} can be modified to:
\begin{gather}
X_{ijt} \sim \mathrm{Bin}(N_{it}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{Normal}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
where $X_{ijt}$ is the number of bird recorded at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the standard deviation of the random observer effects (also on the logit scale). 
We have implemented random observer effects in the Bayesian
framework using program \textbf{JAGS} \citep[version 3.2.0, see Appendix]{plummer:2003}.
% (version 3.2.0; Plummer 2003)
%with the \textbf{R} package \texttt{rjags} \citep{plummer:2011,R-2012} interface.

@@

The classical approach to computing $[N^+_t]$ is based upon
Empirical Bayes methods [CITE]. 
%Note that neither Empiriis not actually a
%Bayesian method, it is used to estim.
%We take the parameter estimates and plug them back in.... For
%example to obtain the posterior distribution for $[N_{i1}]$
%\begin{equation}
%[\hat{N}_{i1}] = \frac{[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},
%\hat{p}]}{\sum_{N_{i1}=0}^K
%[N_{i1}|\hat{\Lambda}][X_{i1}|N_{i1},\hat{p}]}
%  \label{eq:eb1}
%\end{equation}
%Once we have estimated the posterior distribution for the first
%time period, we can then proceed to estimate the posterior
%distribution for subsequent time periods using recursion.
%\begin{equation}
%  [\hat{N}_{it}] =
%  \frac{[N_{it}|N_{it-1},\hat{\bm{\Theta}}][X_{it}|N_{it},
%\hat{p}]}{\sum_{N_{it}=0}^K [N_{it}|N_{it-1},
%\hat{\Theta}][X_{it}|N_{it},\hat{p}]}
%  \label{eq:eb2}
%\end{equation}
%
%Note that when $X_{it}$ is not observed, including for future
%time periods, the distribution $[X_{it}|N_{it}]$ is taken to be
%uniform such that it has no influence on the posterior. In other words,
%the posterior for years with no observed data is determined entirely
%by the estimated posterior from the previous year and the
%population dynamics parameters.
%
If covariates of the dynamics parameters are present, the problem
becomes much harder because, to fully account for uncertainty,
we need a model for the future values of the covariates as well. When
the covariates are continuous, this requires integrating over their
possible values, which becomes very challenging
computationally. Indeed, simulation methods such as MCMC are the
only viable options when numerous covariates are present, which will
often be the case in many ecological studies.

@@

We estimated the parameters for each simulation using the same
initial abundance (Poisson) and dynamics models as were
simulated, implemented in the unmarked library in \textbf{R}.
When run in a maximum likelihood framework, these models require a
a finite upper bound $N_{max}$,
\citep{royle:2004biom,dail_madsen:2011};
%(Royle 2004, Dail and Madsen 2011);
we used 200.  We report bias of estimates, root
mean squared error, and coverage (percentage of 95\% confidence
intervals for parameters that overlap the true values).

@@

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models for each
species, and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant, autoregressive, Ricker-logistic, Gompertz-logistic, autoregressive +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600 for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and, where appropriate, environmental stochasticity or zero-inflated dynamics.  We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 tuning samples.

@@

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
frequentist and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}
